Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-24 paper.qmd
tighter intro

By *linked and executable code and data*, Peng referred to compiled source code
and runnable scripts. In this paper, I interpret this notion more broadly as the
*computational environment*: the complete set of software required to execute an
analysis. Here too, a continuum exists. At the minimal end, authors might only
name the main software used—say, the [R]{.proglang} programming language
[@r-core-team]. More careful authors might also specify the version of
[R]{.proglang}, or list the additional packages and their versions. Rarely,
however, do authors specify the operating system on which the analysis was
performed, even though differences in operating systems can lead to divergent
results when using the same code and software versions, as shown by
@neupane2019. It is even less common for authors to provide step-by-step
installation instructions for the required software stack, an omission often
driven by institutional constraints. Journals, for instance, can inadvertently
work against reproducibility by imposing strict page or word limits that leave
no room for the necessary technical documentation, effectively discouraging
thoroughness in the name of brevity.

Even when such instructions are given, they often fail across different
platforms or versions of the same platform. This lack of portability not only
hinders reproducibility but also complicates everyday research workflows.
Researchers working across multiple machines must be able to recreate the same
environment consistently, and collaborators must share identical computational
setups to avoid inconsistencies.

@@

This paper focuses specifically on two critical but often overlooked aspects of
reproducibility: *computational environment management* and *workflow
orchestration*. I present a comprehensive framework that addresses both
challenges using the [Nix]{.pkg} package manager [@dolstra2004nix], making it
accessible to researchers through two [R]{.proglang} packages: [rix]{.pkg} and
[rixpress]{.pkg}. Before introducing these packages, I survey existing tools and
their limitations to contextualise my contribution.

A range of tools now exist to help researchers approach the gold standard of
full reproducibility, or to consistently deploy the same development environment
across multiple machines. Let us first consider the most basic step in this
process: listing the software used. In [R]{.proglang}, the `sessionInfo()`
function provides a concise summary of the software environment, including the R
version, platform details, and all loaded packages. Its output can be saved to a
file and included as part of a study's reproducibility record. Below is an
example output from `sessionInfo()`:

@@

When an author includes this information, others attempting to reproduce the
study (future readers, collaborators, or even the author at a later time) can
easily see which version of [R]{.proglang} and which packages (with their
versions) were used. However, reproducing the environment still requires
manually installing the correct package versions—a process that becomes
particularly difficult when packages depend on system-level libraries. For
example, the [sf]{.pkg} package requires [GDAL]{.pkg}, [GEOS]{.pkg}, and
[PROJ]{.pkg} system libraries. Installing and configuring these dependencies
varies substantially across operating systems and can be prohibitively difficult
on some platforms.

A more robust approach than simply listing package versions is to use tools that
help data scientists define per-project environments. This is common practice
for [Python]{.proglang} users, who have many tools at their disposal, such as
the built-in *virtual environments* module, but also sophisticated third-party
tools. One particularly popular tool is [uv]{.pkg}, which manages both
project-specific [Python]{.proglang} interpreters (enabling different Python
versions per project) and package libraries using a lock file mechanism.

In the [R]{.proglang} ecosystem, virtual environments are not a built-in
language feature as they are in [Python]{.proglang}, but tools like [renv]{.pkg}
provide equivalent functionality. [renv]{.pkg} captures the project's software
state and writes it to a lock file (called `renv.lock`), which includes the
exact versions of [R]{.proglang} and all required packages. This lockfile serves
as a blueprint for restoring the environment automatically, ensuring that others
can recreate the same setup with minimal effort. Below is an abbreviated example
of an `renv.lock` file:

```{=tex}
\begin{Code}
{
  "R": {
    "Version": "4.2.2",
    "Repositories": [
      {
        "Name": "CRAN",
        "URL": "https://packagemanager.rstudio.com/all/latest"
      }
    ]
  },
  "Packages": {
    "MASS": {
      "Package": "MASS",
      "Version": "7.3-58.1",
      "Source": "Repository",
      "Repository": "CRAN",
      "Hash": "762e1804143a332333c054759f89a706",
      "Requirements": []
    },
    "Matrix": {
      "Package": "Matrix",
      "Version": "1.5-1",
      "Source": "Repository",
      "Repository": "CRAN",
      "Hash": "539dc0c0c05636812f1080f473d2c177",
      "Requirements": ["lattice"]
    }
    ... additional packages omitted ...
  }
}
\end{Code}
```

This lockfile lists each package alongside its version and the repository from
which it was downloaded. Creating it requires only a single command:
`renv::init()`. Others can then reproduce the same package library by running
`renv::restore()`, which installs the exact package versions in an isolated,
project-specific library that does not interfere with the user's global R
library.

However, both [renv]{.pkg} and [uv]{.pkg} have important limitations. While
[uv]{.pkg} manages Python interpreter versions directly, [renv]{.pkg} does not
restore the version of [R]{.proglang} itself—installing the correct R version
must be done separately using tools like [rig]{.pkg} [@rlib2023]. More
critically, neither [uv]{.pkg} nor [renv]{.pkg} handle system-level
dependencies. If [sf]{.pkg} requires [GDAL]{.pkg} version 3.0 but the system has
version 2.4 installed, neither tool can resolve this conflict. Users must
manually install system libraries, and the process differs across operating
systems. Further details are discussed in the [renv]{.pkg} documentation^[See
https://rstudio.github.io/renv/articles/renv.html.].

Pre-compiled binary packages (Python wheels or R binaries from Posit Package
Manager) can simplify installation by avoiding compilation, but they do not
eliminate the need for system libraries that would be required at runtime—these
must still be installed separately.

Other [R]{.proglang} packages provide similar functionality with different
trade-offs. [groundhog]{.pkg} [@simonsohn2023] and [rang]{.pkg} [@chan2023]
enable date-based package installation, retrieving packages as they existed on
CRAN at a specified date. The Posit Package Manager offers another approach
through dated CRAN snapshots. However, like [renv]{.pkg}, these tools focus on
[R]{.proglang} packages and do not address system-level dependencies or
[R]{.proglang} version management.

These tools represent significant progress in reproducibility within their
respective ecosystems, but they share a common limitation: none addresses
system-level dependencies or supports polyglot environments. A project using
[R]{.proglang}, [Python]{.proglang} [@python3], and system tools like
[Quarto]{.pkg} or \LaTeX requires coordinating multiple package managers
([renv]{.pkg} for R, [uv]{.pkg} for [Python]{.proglang}, manual installation for
system tools), each with its own configuration and potential for conflicts.

The most comprehensive approach to reproducibility is containerisation with
[Docker]{.pkg}. [Docker]{.pkg} packages a *data product* together with all its
dependencies—including the operating system, system libraries, programming
languages, and packages—into a self-contained image. A statistical analysis can
be viewed as such a data product. The steps to create an image are specified in
a `Dockerfile`, which defines the structure and content of the [Docker]{.pkg}
image.

Once the image is built, the analysis can be executed inside a *container*,
which is a running instance of the image. [Docker]{.pkg} containers are typically run
non-interactively, allowing the complete computational environment to be
instantiated and executed with a single command.

The strength of [Docker]{.pkg} lies in its ability to bundle not only the
correct versions of [R]{.proglang} and its packages but also system-level
dependencies. Since [Docker]{.pkg} images are effectively minimal Linux systems,
they naturally include libraries required by packages like [sf]{.pkg}. This
ensures that future replicators have access to the same software environment,
including all necessary system components. Moreover, these images can be easily
shared via registries like Docker Hub, enabling seamless reproducibility across
machines and collaborators.

The Rocker project, introduced by @boettiger2017, provides a collection of
pre-built [Docker]{.pkg} images for the [R]{.proglang} community. These images
come with specific R versions and, in some cases, preinstalled packages, making
them convenient base images for building reproducible analysis environments
without starting from scratch.

Despite its strengths, [Docker]{.pkg} presents several challenges for
statistical computing workflows:

1. Poor interactive support: While [Docker]{.pkg} containers can be modified at
  runtime, changes are lost when the container stops unless explicitly saved.
  Running graphical applications like RStudio Desktop from containers requires
  complex X11 configuration that works reliably only on Linux. Instead,
  web-based IDEs (like RStudio Server) are used, but come with the inherent
  limitations of web-based solutions.

2. Steep learning curve: [Docker]{.pkg} images are minimal Linux systems, so
  writing reliable `Dockerfile`s requires familiarity with Linux system
  administration. To ensure reproducibility, `Dockerfile`s should reference
  specific image *digests* rather than *tags*, but this practice is rarely
  followed. Packages like [dockerfiler]{.pkg} [@dockerfiler] help automate
  `Dockerfile` creation but cannot eliminate the need for basic Linux knowledge.

3. Post-hoc reproducibility: A common workflow is to develop analyses
  interactively using standard R installations, then create a `Dockerfile`
  afterward to enable reproduction. While this achieves post-hoc
  reproducibility, it fails to address the challenge of deploying consistent
  environments across multiple machines *during* development, a critical need
  for collaborative projects.

4. Single-language focus: While [Docker]{.pkg} can support polyglot
  environments, orchestrating them requires manual coordination. There is no
  standard way to specify that one step should run in a [Python]{.proglang}
  environment while another runs in R.

5. Incomplete Reproducibility: Reproducibility should be viewed as a spectrum
   rather than a binary property, as emphasised in *Reproducibility in Software
   Engineering* by @dellaieraMasterThesis2024. [Docker]{.pkg} illustrates this
   spectrum well: it achieves strong run-time reproducibility—consistency across
   systems (space)—but weak build-time reproducibility, which concerns
   consistency across time. As defined in the thesis, build-time reproducibility
   means producing identical artefacts regardless of when or where they are
   built. [Docker]{.pkg} cannot guarantee this, as its reliance on mutable base
   images and non-deterministic commands such as `apt-get update` (or similar
   commands, depending on which base image is used) introduces temporal drift:
   rebuilding the same `Dockerfile` at different times can yield different
   results. Even with pinned base images, the thesis found that container
   checksums varied across builds and machines. Thus, while [Docker]{.pkg}
   provides strong run-time reproducibility, achieving verifiable, deterministic
   artefacts requires pinning image digests, freezing dependencies, and using
   deterministic build systems such as [Nix]{.pkg} for full control over both
   temporal and spatial dimensions [@Malka_2024].

@@

Within [R]{.proglang}, the [targets]{.pkg} package by @landau2021 provides a
modern, declarative approach to workflow management. [targets]{.pkg} tracks
dependencies between code and data, caches intermediate results, and only
recomputes steps affected by changes. This dramatically improves efficiency for
complex analyses.

However, [targets]{.pkg} operates within a single R session. While it can call
[Python]{.proglang} via [reticulate]{.pkg} or execute system commands, all steps
share the same computational environment. For truly polyglot pipelines (where
different steps may require incompatible dependencies or different language
versions) this limitation becomes problematic. One might run [targets]{.pkg}
inside a [Docker]{.pkg} container to ensure reproducibility, but this requires
the entire pipeline to use a single environment.

@mcdermott2021 provides an excellent example of achieving the gold standard of
reproducibility. The accompanying GitHub
repository^[https://github.com/grantmcdermott/skeptic-priors] demonstrates the
state of the art:

- Package versions recorded in an `renv.lock` file
- A `Makefile` automating the full analysis
- A `Dockerfile` providing a complete computational environment

However, achieving this required mastering multiple complex tools: [renv]{.pkg}
for package management, [Docker]{.pkg} for environment containerisation, and
[Make]{.pkg} for workflow orchestration. Each tool has its own syntax, concepts,
and failure modes. The complexity compounds for polyglot projects; for instance,
a [Python]{.proglang} project would require learning virtual environments, and
coordinating [R]{.proglang} and [Python]{.proglang} in a single analysis
requires additional tooling.

What researchers need is a tool that:

1. Manages complete environments including programming languages, packages, and
   system dependencies;
2. Supports multiple languages natively, not through workarounds;
3. Works interactively without the complexity of containers;
4. Provides step-level isolation so different pipeline steps can use
  different environments;
5. Ensures bit-for-bit reproducibility through deterministic builds;
6. Remains simple enough for researchers without systems administration
  expertise.

The [Nix]{.pkg} package manager provides these capabilities. [Nix]{.pkg} ensures
reproducible software installation by deploying *component closures*—packages
bundled with all their dependencies and transitive dependencies
[@dolstra2004nix]. Think of installing a package like packing for a trip.
Traditional package managers are like packing just your clothes, assuming you'll
find a toothbrush and soap at your destination. [Nix]{.pkg} packs everything you
might need—the entire self-contained kit. This is what [Nix]{.pkg} calls a
*component closure*.

This creates self-contained software environments that work
identically across machines and over time. [Nix]{.pkg} can replace
[Docker]{.pkg} for environment isolation, [renv]{.pkg} for package management,
and even [Make]{.pkg} for workflow orchestration—all within a single, unified
framework.

However, [Nix]{.pkg} has a steep learning curve. It is a complex system with its
own purely functional programming language designed to declaratively define how
software is built and configured. While this functional approach ensures
reproducibility, it creates a significant barrier to adoption.

@@

- [rix]{.pkg} generates [Nix]{.pkg} expressions from intuitive R function calls,
eliminating the need to learn the [Nix]{.proglang} language. It handles
environment definition including [R]{.proglang}, [Python]{.proglang},
[Julia]{.proglang} [@bezanson2017julia], system tools, and their dependencies.

- [rixpress]{.pkg} orchestrates polyglot analytical pipelines using [Nix]{.pkg}
as the build engine. Each pipeline step runs in its own hermetically sealed
environment with automatic caching and dependency tracking. This enables true
polyglot workflows where [R]{.proglang}, [Python]{.proglang} and
[Julia]{.proglang} steps coexist with different dependencies.
[Python]{.proglang} user can use [ryxpress]{.pkg}, a port of [rixpress]{.pkg}.

Together, these packages provide what existing tools cannot: a single framework
for managing reproducible, polyglot computational environments and executing
complex analytical workflows with step-level isolation. While [Nix]{.pkg}
remains complex for advanced use cases, [rix]{.pkg} and [rixpress]{.pkg}
abstract this complexity for common research workflows, making deep
reproducibility accessible without requiring systems administration expertise.

The remainder of this paper proceeds as follows. Section 2 introduces the
[Nix]{.pkg} package manager and explains why its functional approach enables
reproducibility. Section 3 presents [rix]{.pkg} and demonstrates environment
definition for various use cases. Section 4 discusses the `rstats-on-nix` fork
of the official package repository and how it addresses practical limitations.
Section 5 introduces [rixpress]{.pkg} and demonstrates polyglot pipeline
orchestration through a complete example. Section 6 concludes with discussion of
appropriate use cases and future directions.

>>>>

By *linked and executable code and data*, Peng referred to compiled source code
and runnable scripts. In this paper, I interpret this notion more broadly as the
*computational environment*: the complete set of software required to execute an
analysis. Here too, a continuum exists. At the minimal end, authors might only
name the main software used—say, the [R]{.proglang} programming language
[@r-core-team]. More careful authors might also specify the version of
[R]{.proglang}, or list the additional packages and their versions. Rarely,
however, do authors specify the operating system on which the analysis was
performed, even though differences in operating systems can lead to divergent
results when using the same code and software versions, as shown by
@neupane2019. It is even less common for authors to provide step-by-step
installation instructions for the required software stack, an omission often
driven by institutional constraints. Journals, for instance, can inadvertently
hinder reproducibility by imposing strict page or word limits that leave no room
for the necessary technical documentation, discouraging thoroughness in the name
of brevity.

Even when such instructions are given, they often fail across different
platforms or versions of the same platform. This lack of portability not only
hinders reproducibility but also complicates everyday research workflows.
Researchers using multiple machines must recreate environments consistently, and
collaborators must share identical computational setups to avoid
inconsistencies.

@@

This paper focuses on two critical but often overlooked aspects of
reproducibility: *computational environment management* and *workflow
orchestration*. I present a comprehensive framework addressing both challenges
using the [Nix]{.pkg} package manager [@dolstra2004nix], making it accessible to
researchers through two [R]{.proglang} packages: [rix]{.pkg} and
[rixpress]{.pkg}. Before introducing these packages, I survey existing tools and
their limitations to contextualise my contribution.

A range of tools now exist to help researchers approach the gold standard of
full reproducibility, or to consistently deploy the same development environment
across multiple machines. Let us first consider the most basic step in this
process: listing the software used. In [R]{.proglang}, the `sessionInfo()`
function provides a concise summary of the software environment, including the R
version, platform details, and all loaded packages. Its output can be saved to a
file and included as part of a study’s reproducibility record. Below is an
example output from `sessionInfo()`:

@@

When an author includes this information, others attempting to reproduce the
study (future readers, collaborators, or even the author at a later time) can
see which version of [R]{.proglang} and which packages (with their versions)
were used. However, reproducing the environment still requires manually
installing the correct package versions—a process that becomes difficult when
packages depend on system-level libraries. For example, the [sf]{.pkg} package
requires [GDAL]{.pkg}, [GEOS]{.pkg}, and [PROJ]{.pkg} libraries. Installing and
configuring these dependencies varies substantially across operating systems and
can be prohibitively difficult on some platforms.

A more robust approach than simply listing package versions is to use tools that
define per-project environments. This is common for [Python]{.proglang} users,
who have many tools at their disposal, such as the built-in *virtual
environments* module, or third-party tools like [uv]{.pkg}, which manages both
project-specific interpreters and package libraries using a lock file mechanism.

In the [R]{.proglang} ecosystem, virtual environments are not built-in as they
are in [Python]{.proglang}, but tools like [renv]{.pkg} provide equivalent
functionality. [renv]{.pkg} captures the project’s software state and writes it
to a lock file (`renv.lock`), which includes the exact versions of
[R]{.proglang} and all required packages. This lockfile serves as a blueprint
for restoring the environment automatically, ensuring others can recreate the
same setup with minimal effort.

However, both [renv]{.pkg} and [uv]{.pkg} have limitations. While [uv]{.pkg}
manages Python interpreter versions directly, [renv]{.pkg} does not restore the
version of [R]{.proglang} itself—installing the correct version must be done
separately using tools like [rig]{.pkg} [@rlib2023]. More critically, neither
handles system-level dependencies. If [sf]{.pkg} requires [GDAL]{.pkg} version
3.0 but the system has version 2.4 installed, neither tool can resolve this
conflict. Users must manually install system libraries, and the process differs
across operating systems.

Pre-compiled binaries simplify installation but do not eliminate the need for
system libraries required at runtime. Other [R]{.proglang} packages such as
[groundhog]{.pkg} [@simonsohn2023] and [rang]{.pkg} [@chan2023] enable
date-based package installation, while the Posit Package Manager provides dated
CRAN snapshots. However, like [renv]{.pkg}, these focus on R packages and do not
address system-level dependencies or R version management.

These tools represent significant progress in reproducibility within their
ecosystems, but they share a common limitation: none handles system-level
dependencies or polyglot environments. A project using [R]{.proglang},
[Python]{.proglang} [@python3], and system tools like [Quarto]{.pkg} or \LaTeX
requires coordinating multiple package managers ([renv]{.pkg} for R, [uv]{.pkg}
for Python, manual installation for system tools), each with its own
configuration and potential conflicts.

The most comprehensive approach to date is containerisation with [Docker]{.pkg}.
[Docker]{.pkg} packages a *data product* together with all its
dependencies—including the operating system, system libraries, programming
languages, and packages—into a self-contained image. Once built, the analysis
can be executed inside a *container*, a running instance of the image.

The strength of [Docker]{.pkg} lies in its ability to bundle not only the
correct versions of [R]{.proglang} and its packages but also system-level
dependencies. Since [Docker]{.pkg} images are minimal Linux systems, they
include the libraries required by packages such as [sf]{.pkg}, ensuring that
future replicators have access to the same environment. Moreover, these images
can be shared via registries like Docker Hub, enabling reproducibility across
systems.

The Rocker project [@boettiger2017] provides pre-built [Docker]{.pkg} images for
the [R]{.proglang} community, offering specific R versions and, in some cases,
preinstalled packages. These serve as convenient bases for reproducible analysis
environments.

Despite its strengths, [Docker]{.pkg} presents several challenges:

1. Poor interactive support: While containers can be modified at runtime,
   changes are lost unless explicitly saved. Running graphical applications like
   RStudio Desktop requires complex X11 configuration that works reliably only
   on Linux. Web-based IDEs such as RStudio Server are alternatives but have
   their own limitations.

2. Steep learning curve: Writing reliable `Dockerfile`s requires familiarity
   with Linux system administration. To ensure reproducibility, `Dockerfile`s
   should reference image *digests* rather than *tags*, but this is rarely done.
   Packages like [dockerfiler]{.pkg} [@dockerfiler] help automate this but do
   not remove the need for basic Linux knowledge.

3. Post-hoc reproducibility: Analyses are often developed interactively and
   containerised only afterward, achieving post-hoc reproducibility rather than
   consistent environments during development.

4. Single-language focus: While [Docker]{.pkg} can support polyglot setups,
   orchestrating them requires manual coordination. There is no standard way to
   specify that one step runs in [Python]{.proglang} and another in [R]{.proglang}.

5. Incomplete reproducibility: Reproducibility is best viewed as a spectrum
   rather than a binary property [@dellaieraMasterThesis2024]. [Docker]{.pkg}
   achieves strong run-time reproducibility—consistency across systems
   (space)—but weak build-time reproducibility, or consistency over time.
   Mutable base images and non-deterministic commands such as `apt-get update`
   introduce temporal drift: rebuilding the same `Dockerfile` at different times
   can yield different results. Even with pinned base images, container
   checksums can vary across builds and machines. Thus, while [Docker]{.pkg}
   provides strong run-time reproducibility, achieving deterministic artefacts
   requires pinning image digests, freezing dependencies, and using
   deterministic build systems such as [Nix]{.pkg} for full control over both
   time and space [@Malka_2024].

@@

Within [R]{.proglang}, the [targets]{.pkg} package [@landau2021] provides a
modern, declarative approach to workflow management. It tracks dependencies,
caches results, and only recomputes affected steps, improving efficiency for
complex analyses. However, [targets]{.pkg} operates within a single
[R]{.proglang} session. For truly polyglot pipelines—where steps may require
incompatible dependencies or different language versions—this becomes
problematic. Running [targets]{.pkg} inside a [Docker]{.pkg} container ensures
reproducibility but forces the entire pipeline to share one environment.

@mcdermott2021 provides an excellent example of best practice. The accompanying
repository^[See
[https://github.com/grantmcdermott/skeptic-priors](https://github.com/grantmcdermott/skeptic-priors)]
demonstrates:

- Package versions recorded in an `renv.lock` file;
- A `Makefile` automating the full analysis;
- A `Dockerfile` providing the computational environment.

Achieving this required mastering multiple complex tools: [renv]{.pkg} for
package management, [Docker]{.pkg} for containerisation, and [Make]{.pkg} for
orchestration. The complexity increases for polyglot projects, which must manage
multiple ecosystems simultaneously.

Researchers therefore need a tool that:

1. Manages complete environments including languages, packages, and system
   dependencies;
2. Supports multiple languages natively;
3. Works interactively without container complexity;
4. Provides step-level isolation for different environments;
5. Ensures bit-for-bit reproducibility through deterministic builds;
6. Remains simple enough for researchers without systems administration
   expertise.

The [Nix]{.pkg} package manager provides these capabilities. It ensures
reproducible installation by deploying *component closures*—packages bundled
with all their direct and transitive dependencies [@dolstra2004nix]. Think of
installing a package as packing for a trip: traditional package managers assume
you’ll find essentials at your destination, while [Nix]{.pkg} packs everything
you need. This produces self-contained software environments that behave
identically across machines and over time.

[Nix]{.pkg} can replace [Docker]{.pkg} for isolation, [renv]{.pkg} for package
management, and even [Make]{.pkg} for orchestration—all within a single, unified
framework. However, it has a steep learning curve, with its own functional
language for declaratively defining software builds and configurations. While
this ensures reproducibility, it also creates a barrier to adoption.

@@

- [rix]{.pkg} generates [Nix]{.pkg} expressions from intuitive R function calls,
  eliminating the need to learn the [Nix]{.proglang} language. It handles
  environment definitions including [R]{.proglang}, [Python]{.proglang},
  [Julia]{.proglang} [@bezanson2017julia], system tools, and dependencies.

- [rixpress]{.pkg} orchestrates polyglot analytical pipelines using [Nix]{.pkg}
  as the build engine. Each step runs in its own hermetic environment with
  automatic caching and dependency tracking, enabling true polyglot workflows
  across R, Python, and Julia. [Python]{.proglang} users can use
  [ryxpress]{.pkg}, a direct port.

Together, these packages provide a single framework for managing reproducible,
polyglot environments and executing complex workflows with step-level isolation.
While [Nix]{.pkg} remains complex for advanced use, [rix]{.pkg} and
[rixpress]{.pkg} abstract this complexity, making deep reproducibility
accessible without systems administration expertise.

The remainder of this paper proceeds as follows. Section 2 introduces
[Nix]{.pkg} and explains how its functional model enables reproducibility.
Section 3 presents [rix]{.pkg} and demonstrates environment definition. Section
4 discusses the `rstats-on-nix` fork of the package repository. Section 5
introduces [rixpress]{.pkg} and demonstrates polyglot pipeline orchestration.
Section 6 concludes with discussion of future directions.
